Physcraper: Body

### 2020-09-01 2-body.Rmd
word limit includes references :sob:

Phylogenetic estimates of evolutionary relationships capture the shared history of living and extinct organisms and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with new DNA sequence data, geographical location, and other metadata in a reproducible and continuous manner is possible by automating connections between biological databases. Here, we introduce Physcraper, a tool to build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope, by using sequence data from public biological databases to update a starting tree and single locus alignments.

Automatic integration of data in a phylogenetic context is challenged by the prevalence of taxonomic idiosyncrasies across biological databases. Taxonomic name standardization represents a step toward tackling this issue, but requires a unified system. The Open Tree of Life project (OpenTree)
<!-- [project](https://opentreeoflife.github.io/)  -->
has constructed a comprehensive and digitally-available tree of life by synthesizing published phylogenies along with taxonomic data. Currently, OpenTree's "synthetic" tree comprises 2.3 million tips, of which around 90,000 are supported by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases [@rees2017automated], including the USA National Center for Biodiversity Information (NCBI) molecular database GenBank [@benson2000genbank; @wheeler2000database], the Global Biodiversity Information Facility [GBIF; @secretariat2017gbif], the World Register of Marine Species (WoRMS), and other resources. The OpenTree unified taxonomy represents a key resource for connecting data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multiple single locus data sets taking into account the multispecies coalescent model are considered the gold standard for inferring species relationships [@song2012resolving].
Genomics has, and will continue to, revolutionize phylogenetic inference.
Yet, different research questions call for different genomic sequencing approaches (e.g., whole genomes, microsatellites, ultra-conserved elements), which produce largely non-overlapping genomic data sets across taxa, creating challenges in wide scale phylogenetic reconstruction.
<!-- Also, as @andermann2020guide point out, "assembling full genomes is often unnecessary for phylogenomic studies if the main goal is to retrieve an appropriate number of phylogenetically informative characters from several independent and single-copy genetic markers [@jones2016targeted]." -->
While phylogenomics ameliorates the problem of non-overlapping genomic data sets by focusing on targeted capture of informative characters from independent and single-copy genetic markers [@andermann2020guide; @jones2016targeted], decades of single locus sequencing have already generated massive amounts of homologous DNA data sets that can be used for phylogenetic reconstruction at many scales.
Even in phylogenomic reconstructions, thoughtfully curated markers and alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

@@

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a public source of valuable expert knowledge. Linking published alignments with molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life. The phylogenies associated with TreeBASE alignments have been integrated to the OpenTree's datastore, Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a framework linking trees with the alignment that generated them.

@@

The DNA sequence search is performed with the Basic Local Alignment Search Tool, BLAST [@altschul1990basic; @altschul1997gapped], either on the GenBank remote database or in a GenBank local database set up by the user. It is constrained to a taxonomic group in the NCBI taxonomy, defined as the "search taxon". Users can arbitrarily define a search taxon that is either a more or a less inclusive
clade relative to the ingroup of the input tree.
Otherwise, the search taxon is automatically identified using the OpenTree [API](<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>) [@rees2017automated], as the Most Recent Common Ancestral Taxon (MRCAT) of the ingroup taxa in the input tree, i.e., the MRCA that is also a named clade in the NCBI taxonomy (Fig. \@ref(fig:framework)).
  <!-- FIGURE RECOMMENDED: Figure \@ref(fig:search)  -->
The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.

@@

Second, a MUSCLE profile alignment is performed, in which the original alignment is used as a template to align the new sequences. This ensures that the final alignment
follows the homology criteria established by the original alignment.
The final alignment is not further processed by Physcraper. It is recommended that the alignment is checked by the user, by eye followed by manual refinement, or using a tool for automatic alignment processing [e.g., GBlocks; @castresana2000selection; @castresana2002gblocks].
Users may also use Physcraper to only gather new GenBank sequences, to then apply their own preferred alignment and phylogenetic inference methods.

@@

Physcraper's final result is an updated phylogenetic hypothesis for the search taxon based on the data in the input alignment.
The updated tree can be compared with the original tree using an automated conlfict analysis which calculates Robinson Foulds weighted and unweighted metrics are estimated using Dendropy functions [@sukumaran2010dendropy].
Conflict analysis also performs a node by node comparison between the synthetic OpenTree and the original and updated tree individually, using OpenTree's conflict API [@redelings2017supertree].
For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined. A suggested default rooting based on OpenTree's taxonomy is implemented for now. This approach uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the root may no longer be appropriate. Automatic identification of a phylogenetic tree root is indeed a difficult problem that has not been solved yet. The best way right now is for users to define the outgroup directly on the updated tree as part of the conflict analysis, so trees are accurately rooted.
<!-- *It would be a nice addition to have users give the output of the input tree as an argument at some point, or maybe we could add a super outgroup at random based on the search taxon* -->

@@

Data repositories designed to preserve and democratize access to biological data
constitute an essential resource for evolutionary estimation. While data keep accumulating,
our ability to reanalyze and incorporate new data with available knowledge is being challenged.

Various phylogenetic pipelines have been designed to make evolutionary sense of the vast amount of public molecular data available in different ways.
Pipelines like Phylota [@sanderson2008phylota], PyPHLAWD [@smith2009mega], and SUPERSMART [@antonelli2017toward]
focus mainly on generating full trees *de novo* (i.e., inferring phylogenetic relationships from a newly generated homology hypothesis, as opposed to e.g., supertrees, that are generated by assembling previous phylogenetic estimates) for regions of the Tree of
Life that have no phylogenetic assessment yet in published studies.
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of pre-existing
phylogenetic knowledge available in tree repositories to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline [@izquierdo2014pumper] also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to conduct a comparison analysis to Physcraper.
<!-- **MTH: add some more details about what aspect of the PUMPER install failed** -->

Physcraper on its own generates individual gene trees, which do not capture the full complexity of species' evolutionary history [@song2012resolving]. However, using Physcraper it is straightforward to gather alignments and gene trees for multiple loci for a taxonomic group of interest, which can be used as inputs for downstream tools to perform coalescent analyses and assess species tree relationships, (e.g. ASTRAL [@mirarab2014astral], BEAST2 [@bouckaert2019beast], SVD Quartets [@chifman2014quartet]).

Physcraper is novel in that it links molecular and tree data repositories.
It can also link inferences to various biological databases, such as GBIF, the Paleobiology Database, and others, by leveraging the integrated taxonomy from OpenTree [@rees2017automated], a resource that is contributing to the goal described by [@wilson2003encyclopedia] of "exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science".

Physcraper in conjuction with OpenTree can be used to rapidly (in a matter of hours)
to address challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically [@webb2010biodiversity],
obtaining trees for ecophylogenetic studies [@helmus2012phylogenetic],
systematizing molecular (and other) databases, i.e., curating taxonomic assignations [@san2010molecular],
and generating custom species trees for ecological and evolutionary downstream analyses [@stoltzfus2013phylotastic].

@@

\newpage
# References

>>>>

Phylogenies capture the shared history of living and extinct organisms and provide key evolutionary context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary studies, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenies with new DNA sequence data, geographical location, and other metadata in a reproducible and continuous manner is possible by automating connections between biological databases. Here, we introduce Physcraper, a tool to build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope, by using sequence data from public biological databases to update a starting tree and single locus alignments.

Automatic integration of data in a phylogenetic context is challenged by the prevalence of taxonomic idiosyncrasies across biological databases. Taxonomic name standardization represents a step toward tackling this issue, but requires a unified system. The Open Tree of Life project (OpenTree)
<!-- [project](https://opentreeoflife.github.io/)  -->
has constructed a comprehensive tree of life by synthesizing published phylogenies along with taxonomic data. OpenTree's "synthetic" tree comprises 2.3 million tips, of which around 90,000 are supported by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on taxonomy. To achieve this, OpenTree unifies taxonomy from various databases [@rees2017automated], including the USA National Center for Biodiversity Information (NCBI) molecular database GenBank [@benson2000genbank; @wheeler2000database], the Global Biodiversity Information Facility [GBIF; @secretariat2017gbif], the World Register of Marine Species (WoRMS), and other resources. The OpenTree unified taxonomy represents a key resource for connecting data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses.
Species tree reconstructions from multiple single locus data sets taking into account the multispecies coalescent model are considered the gold standard for inferring species relationships [@song2012resolving].
Genomics has, and will continue to, revolutionize phylogenetic inference.
Yet, different research questions call for different genomic sequencing approaches (e.g., whole genomes, microsatellites, ultra-conserved elements), which produce largely non-overlapping genomic data sets across taxa, creating challenges in wide scale phylogenetic reconstruction.
<!-- Also, as @andermann2020guide point out, "assembling full genomes is often unnecessary for phylogenomic studies if the main goal is to retrieve an appropriate number of phylogenetically informative characters from several independent and single-copy genetic markers [@jones2016targeted]." -->
While phylogenomics ameliorates the problem of non-overlapping genomic data sets by focusing on targeted capture of informative characters from independent and single-copy genetic markers [@andermann2020guide; @jones2016targeted], decades of single locus sequencing have generated massive amounts of homologous DNA data sets that can be used for phylogenetic reconstruction at many scales.
Even in phylogenomic reconstructions, thoughtfully curated markers and alignments can improve phylogenetic reconstructions [@fragoso2017pilot].

@@

A way to incorporate the benefits from massive amounts of newly released molecular data and fine-grained curation from human experts, is to rely on manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing a public source of valuable expert knowledge. Linking published alignments with molecular data that has not yet been included in any public phylogenetic estimate, has the potential to accelerate the enrichment and updating of phylogenetic relationships in many regions of the tree of life.
<!-- The phylogenies associated with TreeBASE alignments have been integrated to the OpenTree's datastore, Phylesystem [@mctavish2015phylesystem], and metadata linking them to their corresponding alignment repository is available, providing a framework linking trees with the alignment that generated them. -->

@@

The DNA sequence search is performed with the Basic Local Alignment Search Tool, BLAST [@altschul1990basic], either on the GenBank remote database or in a GenBank local database set up by the user. It is constrained to a taxonomic group in the NCBI taxonomy, defined as the "search taxon". Users can arbitrarily define a search taxon that is either a more or a less inclusive
clade relative to the ingroup of the input tree.
Otherwise, the search taxon is automatically identified using the OpenTree [API](<https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca>) [@rees2017automated], as the Most Recent Common Ancestral Taxon (MRCAT) of the ingroup taxa in the input tree, i.e., the MRCA that is also a named clade in the NCBI taxonomy (Fig. \@ref(fig:framework)).
  <!-- FIGURE RECOMMENDED: Figure \@ref(fig:search)  -->
The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.

@@

Second, a MUSCLE profile alignment is performed, in which the original alignment is used as a template to align the new sequences. This ensures that the final alignment
follows the homology criteria established by the original alignment.
The final alignment is not further processed by Physcraper. It is recommended that the alignment is checked by the user, by eye followed by manual refinement, or using a tool for automatic alignment processing [e.g., GBlocks; @castresana2000selection].
Users may also use Physcraper to only gather new GenBank sequences, to then apply their own preferred alignment and phylogenetic inference methods.

@@

Physcraper's main result is an updated phylogenetic hypothesis for the search taxon.
The updated tree can be compared with the original tree using an automated conflict analysis which calculates Robinson Foulds weighted and unweighted metrics using Dendropy  [@sukumaran2010dendropy], and performs a node by node comparison between the synthetic OpenTree and the original and updated tree individually, using OpenTree's conflict API [@redelings2017supertree].
For the conflict analysis to be meaningful, the root of the tree needs to be accurately defined. A default rooting based on OpenTree's taxonomy is implemented for now. It uses the taxon labels for all the tips in the updated tree, pulls an inferred subtree from OpenTree's taxonomy and then applies the same rooting to the inferred updated tree. However, if the updated tree changes expectations from taxonomy, the rooting may no longer be appropriate. Automatic identification of a phylogenetic tree root is a difficult problem that has not been solved yet. The best way right now is for users to define the outgroup directly on the updated tree as part of the conflict analysis, so trees are accurately rooted.
<!-- *It would be a nice addition to have users give the output of the input tree as an argument at some point, or maybe we could add a super outgroup at random based on the search taxon* -->

@@

Data repositories designed to preserve and democratize access to biological data
are essential resources for evolutionary estimation. While data keep accumulating,
our ability to reanalyze and incorporate new data with available knowledge is being challenged.

While phylogenetic pipelines designed to make evolutionary sense of the vast amount of public molecular data are available (e.g., Phylota [@sanderson2008phylota], PyPHLAWD [@smith2009mega], SUPERSMART [@antonelli2017toward]), they focus on generating full trees *de novo* (i.e., inferring phylogenetic relationships from a newly generated homology hypothesis, as opposed to e.g., supertrees, that are generated by assembling previous phylogenetic estimates).
While Physcraper does not generate phylogenies *de novo* in a traditional sense,
it successfully generates new phylogenetic knowledge, revealing the potential of
phylogenetic knowledge published in databases to facilitate phylogenetic placement of public molecular data.
The PUMPER pipeline [@izquierdo2014pumper] also uses the concept of updating
pre-existing alignments to incorporate public molecular data into phylogenies. Unfortunately, installation of the tool was unsuccesful following instructions from the author, and we were unable to benchmark a comparison.
<!-- **MTH: add some more details about what aspect of the PUMPER install failed** -->

Physcraper generates individual gene trees, which fail to capture the complexity of species' evolutionary history [@song2012resolving]. Yet, Physcraper makes it straightforward to gather alignments and gene trees for multiple loci from a taxonomic group of interest, which can then be used by any tool that performs coalescent analyses to generate species trees, e.g., ASTRAL [@mirarab2014astral], BEAST2 [@bouckaert2019beast], SVD Quartets [@chifman2014quartet]).

Physcraper not only links molecular and tree databases, it can also link phylogenies to various biological databases, such as GBIF, the Paleobiology Database, and others, by leveraging the integrated taxonomy from OpenTree [@rees2017automated].
<!-- a resource that is contributing to achieve the goal of "exploring and analysing biodiversity at an accelerated pace, and returning systematics into the mainstream of science" [@wilson2003encyclopedia]. -->

Physcraper in conjuction with OpenTree can be used to rapidly (in a matter of hours)
address challenges overarching both fields of ecology and evolution, such as
placing newly discovered species phylogenetically [@webb2010biodiversity],
obtaining trees for ecophylogenetic studies [@helmus2012phylogenetic],
systematizing molecular (and other) databases, i.e., curating taxonomic assignations [@san2010molecular],
and generating custom species trees for ecological and evolutionary downstream analyses [@stoltzfus2013phylotastic].

@@

# References
